Bacon: A GPU Programming Language With Just in Time Specialization

### 2011-05-30 doc/Paper/Paper.tex
Paper tweaks.

This paper describes Bacon, a data-parallel programming system
targetting OpenCL-compatible graphics processors. This system is built
upon the existing OpenCL standard in order to make it easier for
programmers to write high performance kernels for GPU accelerated
applications. The OpenCL C syntax is extended into a new language,
Bacon C, intended to make development significantly more convienient
and enabling pre-optimizations based on just-in-time specialization as
this code is compiled via OpenCL at runtime.

@@

The use of Graphics Processing Units (GPUs) for general purpose
parallel computing has become increasingly feasible over the last few
years. In response to the platform specific programming solutions from
Nvidia and Microsoft, Apple developed the OpenCL standard as an open and
cross platform programming interface to this hardware.

@@

OpenCL C uses the syntax of C99 and provides a set of built in data
types and functions that expose the numeric computatation capabilities
common to modern GPU devices. The specification explicitly disallows
the use of various C99 functionality that is not supported by GPU
hardware, including function pointers, recursion, and any sort of
dynamic memory allocation or array sizing.

Rather than providing a stand-alone program to compile OpenCL C
kernels, the OpenCL C and C++ APIs give developers the nessisary
pieces to build a compiler into their host program. This allows OpenCL
programs to be portable across different hardware archetectures by
delaying compilation until runtime when the target GPU device is
known, but requires each developer to write quite a bit of code to
load the source code and perform various other bookkeeping.

Bacon was developed to try to improve upon the OpenCL programming
experience and make it easier to write high performance programs for
GPU hardware. It does this by extending the OpenCL C syntax into a new
language called Bacon C, and by performing some pre-optimizations as
the Bacon C code is compiled to OpenCL C.

The key optimization performed by the Bacon compiler is just in time
specialization. Rather than performing When a kernel is called, a
specialzied version of the kernel is generated with some of its
variable arguments replaced with compile-time constants. This allows
for a variety of agressive optimizatons to be performed by the OpenCL
compiler. Further, it allows for some variable sized array
declarations to be turned into constant sized arrays that are allowed
by OpenCL.

We evaluate these improvements below using matrix multiplication as a
test case. Both simpler code and improved performance are
demonstrated.

\section{Previous Work}

@@

@@

    float sum = 0.0;
    assert(aa.cols == bb.rows, 
        "Matrixes must have compatible dimensions.");
    for (int kk = 0; kk < aa.cols; ++kk) {
        sum += aa[$row, kk] * bb[kk, $col];
    }
    cc[$y, $x] = sum;
    return cc;
}
\end{verbatim}
\caption{Naive Matrix Multiplication in Bacon C}\label{mmk}
\end{listing}

@@

Each BODY includes a @range declaration that specifes the range it
will be executed over in parallel. Within the BODY, the current
position in that range is held in special variables named {\tt\$row},
{\tt\$col}, and {\tt\$dep} for the first, second, and third dimension
respectively.

Parameterized types for 1D, 2D and 3D arrays are provided
natively. The types are parameterized using C++-style angle bracket
syntax. Both declarations and element access use a comma separated
list of dimensions in square brackets. The dimensions of these arrays
can be accessed using struct-style dot notation.

@@

The generated C++ wrapper provides a C++ function with the kernel's
type signature that can be called from the user's application. When
this functions is called, the runtime library loads the AST and walks
it to generate the specialized OpenCL code. Optimizatons are performed
at code generation time without any intermediate pass that transforms
the representation.

@@

The implemetnation of Bacon is available publically under an open
source license. The current version can be downloaded from the public
git repository\footnote{http://code.ferrus.net/compilers/bacon.git}.

@@

Testing was performed on an ATI Radeon HD 5830 GPU. This is a cut down
version of AMD's current top of the line GPU, the Radeon HD 6970. The
5830 sells for around \$140 at the time of this writing compared to
\$340 for the 6970, while providing over 60\% of the theoretical
compute power. A similar Nvidia card was purchased for comparison
testing, but the comparsion was not performed due to implementation
incompatibilities that we have not yet resolved.

@@

\begin{description}
  \item[Bacon - Naive] Shown in Listing~\ref{mmk}.
  \item[OpenCL - Naive] An equivilent OpenCL C implementation.
  \item[Bacon - Blocked] Shown in listing~\ref{bmmk}.
  \item[OpenCL - Hand Vectorized] Hand unrolled using vector types to
    compute 4x4 blocks. Based on a sample from the AMD OpenCL SDK.
\end{description}

Timings were taken multiplying 4096x4096 matrices. Each test was
performed five times and the average result was taken. The times were
very consistent; most tests had a coefficient of variation under one
percent. The speedups over the naive OpenCL implementationa are shown
in Figure~\ref{unroll}.

@@

The Bacon system could benefit from a number of extensions to improve
its performance, generality, and utility to developers. One obvious
area of improvement would be to automatically determine the best
factor for loop unrolling rather than expecting it to be tuned
manually. The unrolling analysis could also be extended to the
implicit parallel ``loops'' created by the range of the computation;
optimally, it should be possible to generate kernels like the blocked
matrix mulitplication shown from the naive matrix multiplication
kernel. Another simple improvement would be to perform strip mining
(Wolfe \cite{wolfe:1996}, section 9.8) and generate explicit vector
code for loops.

Just in time specialization as a method to improve performance on
parallel computations seems very promising in general, but there are
some limitations imposed by the use of OpenCL as a compiler target. An
implementation of this technique targetting a parallel processor at a
lower level would allow for more flexibiltiy and much faster
specialization times.

>>>>

This paper describes Bacon, a data-parallel programming system
targeting OpenCL-compatible graphics processors. This system is built
upon the existing OpenCL standard in order to make it easier for
programmers to write high performance kernels for GPU accelerated
applications. The OpenCL C syntax is extended into a new language,
Bacon C, intended to make development significantly more convenient
and enabling pre-optimizations based on just-in-time specialization as
this code is compiled via OpenCL at runtime.

@@

The use of Graphics Processing Units (GPUs) for general purpose
parallel computing has become increasingly feasible over the last few
years. In response to the platform specific programming solutions from
Nvidia and Microsoft, Apple developed the OpenCL standard as an open
and cross platform programming interface to this hardware. This
standard has since been implemented by a number of major vendors
including AMD, Nvidia, Intel, and IBM.

@@

OpenCL C uses the syntax of C99 and provides a set of built in data
types and functions that expose the numeric computation capabilities
common to modern GPU devices. The specification explicitly disallows
the use of various C99 functionality that is not supported by GPU
hardware, including function pointers, recursion, and any sort of
dynamic memory allocation or array sizing.

Rather than providing a stand-alone program to compile OpenCL C
kernels, the OpenCL C and C++ APIs give developers the pieces
necessary to build a compiler into their host program. This allows OpenCL
programs to be portable across different hardware by architectures
delaying compilation until runtime when the target GPU device is
known, but requires each developer to write quite a bit of code to
load the source code and perform various other bookkeeping activities.

Bacon was developed to improve upon the OpenCL programming
experience and make it easier to write high performance programs for
GPU hardware. It does this by extending the OpenCL C syntax into a new
language called Bacon C, and by performing pre-optimizations as
the Bacon C code is compiled to OpenCL C.

The key optimization performed by the Bacon compiler is just in time
specialization. When a Bacon kernel is written, some integer arguments
can be marked as specialization-time constants. When a kernel is first
called with a given set of values for those arguments a specialized
version of that kernel is generated with those variable arguments
replaced with constant values. This allows for a variety of
optimizations be performed by the OpenCL compiler. Further, it allows
for any variable sized array declarations that depend only on these
specialization-time constants to be turned into constant sized arrays
that are allowed by OpenCL.

These improvements are evaluated using matrix multiplication as a test
case. Both simpler code and improved performance are demonstrated.

\section{Previous and Similar Work}

\subsection{Partial Evaluation}

@@

\subsection{OpenCL Libraries and Front-Ends}

Bacon is not the first attempt to provide an improved programmer
experience for OpenCL.

@@

    float sum = 0.0;
    assert(aa.cols == bb.rows, 
        "Matrices must have compatible dimensions.");
    for (int kk = 0; kk < aa.cols; ++kk) {
        sum += aa[$row, kk] * bb[kk, $col];
    }
    cc[$y, $x] = sum;
    return cc;
}
\end{verbatim}
\caption{Naive Matrix Multiplication in Bacon C}\label{mmk}
\end{listing}

@@

Each BODY includes a @range declaration that specifies the range it
will be executed over in parallel. Within the BODY, the current
position in that range is held in special variables named {\tt\$row},
{\tt\$col}, and {\tt\$dep} for the first, second, and third dimension
respectively.

Parametrized types for 1D, 2D and 3D arrays are provided
natively. The types are parametrized using C++-style angle bracket
syntax. Both declarations and element access use a comma separated
list of dimensions in square brackets. The dimensions of these arrays
can be accessed using struct-style dot notation.

@@

The generated C++ wrapper provides a C++ function with the kernel's
type signature that can be called from the user's application. When
this functions is called, the runtime library loads the AST and walks
it to generate the specialized OpenCL code. Optimizations are performed
at code generation time without any intermediate pass that transforms
the representation.

@@

The implementation of Bacon is available publicly under an open
source license. The current version can be downloaded from the public
git repository\footnote{http://code.ferrus.net/compilers/bacon.git}.

@@

Testing was performed on an ATI Radeon HD 5830 GPU. This is a cut down
version of AMD's current top of the line GPU, the Radeon HD 6970. The
5830 sells for around \$140 at the time of this writing compared to
\$340 for the 6970, while providing over 60\% of the theoretical
compute power. A similar Nvidia card was purchased for comparison
testing, but the comparison was not performed due to implementation
incompatibilities that we have not yet resolved.

@@

\begin{description}
  \item[Bacon - Naive] Shown in Listing~\ref{mmk}.
  \item[OpenCL - Naive] An equivalent OpenCL C implementation.
  \item[Bacon - Blocked] Shown in listing~\ref{bmmk}.
  \item[OpenCL - Hand Vectorized] Hand unrolled using vector types to
    compute 4x4 blocks. Based on a sample from the AMD OpenCL SDK.
\end{description}

Timings were taken multiplying 4096x4096 matrices. Each test was
performed five times and the average result was taken. The times were
very consistent; most tests had a coefficient of variation under one
percent. The speedups over the naive OpenCL implementation are shown
in Figure~\ref{unroll}.

@@

The Bacon system could benefit from a number of extensions to improve
its performance, generality, and utility to developers. One obvious
area of improvement would be to automatically determine the best
factor for loop unrolling rather than expecting it to be tuned
manually. The unrolling analysis could also be extended to the
implicit parallel ``loops'' created by the range of the computation;
optimally, it should be possible to generate kernels like the blocked
matrix multiplication shown from the naive matrix multiplication
kernel. Another simple improvement would be to perform strip mining
(Wolfe \cite{wolfe:1996}, section 9.8) and generate explicit vector
code for loops.

Just in time specialization as a method to improve performance on
parallel computations seems very promising in general, but there are
some limitations imposed by the use of OpenCL as a compiler target. An
implementation of this technique targeting a parallel processor at a
lower level would allow for more flexibility and much faster
specialization times.
